---
title: "Assignment 5 (R: Loops, functions & data structures)"
subtitle: "R: Loops, functions, atomic vectors and matrices"
format:
  html:
    toc: true
self-contained: true
link-external-newwindow: true
editor_options: 
  chunk_output_type: console
---

## Instructions {-}

1. You may talk to a friend, discuss the questions and potential directions for solving them. However, you need to write your own solutions and code separately, and not as a group activity. 

2. Do not write your name on the assignment.

3. Make R code chunks to insert code and type your answer outside the code chunks. Ensure that the solution is written neatly enough to understand and grade.

4. Render the file as HTML to submit.

5. There are 5 points for clealiness and organization. The code should be commented and clearly written with intuitive variable names. For example, use variable names such as number_input, factor, hours, instead of a,b,xyz, etc.

6. The assignment is worth 100 points, and is due on **14th November 2022 at 11:59 pm**. 

## Streaming platforms

The streaming platforms used by the students of two sections of the STAT303-1 Fall 2022 class are give below as two atomic vectors - `streaming_platform_section1` and `streaming_platform_section2`

```{r}
streaming_platform_section1<-c('Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'HBO Max;Disney+;Paramount Plus, Peacock', 'Netflix', 'Netflix;Hulu;Disney+', 'Netflix;Amazon Prime;HBO Max', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu', 'None', 'Amazon Prime', 'Netflix;Hulu;Amazon Prime', 'Amazon Prime', 'Netflix', 'Netflix', 'Netflix', 'none', 'Netflix;Amazon Prime', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Hulu;HBO Max', 'Netflix;Disney+', 'Netflix', 'Netflix;HBO Max', 'Netflix;Hulu', 'Netflix;Amazon Prime;HBO Max;Disney+', 'Netflix;Disney+', 'Netflix;Disney+', 'Netflix;HBO Max;Disney+', 'Netflix', 'Netflix;Amazon Prime', 'Netflix;Disney+', 'HBO Max;Disney+', 'Netflix;Amazon Prime', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+;Shudder, Apple TV+', 'Netflix;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime', 'Hulu;Amazon Prime', 'Netflix;Hulu', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu', 'Netflix;Hulu;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime', 'Netflix;Hulu;Amazon Prime', 'Netflix;Amazon Prime', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Amazon Prime;HBO Max', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime', 'Netflix;Hulu;HBO Max', 'Netflix', 'Netflix', 'Netflix;Hulu;Amazon Prime;HBO Max', 'Netflix;Hulu;Amazon Prime;HBO Max', 'Netflix;Amazon Prime', 'Netflix', 'Disney+', 'Netflix;Amazon Prime;HBO Max', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+;Peacock, Paramount+', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Amazon Prime', 'Netflix;HBO Max', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix', 'Netflix;Amazon Prime;HBO Max', 'Netflix', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix', 'Netflix;Amazon Prime;HBO Max', 'Netflix;Hulu;Amazon Prime;HBO Max', 'Netflix', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max', 'Netflix;Disney+', 'Amazon Prime;Disney+', 'Netflix;Amazon Prime', 'Netflix;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max', 'Netflix;Hulu;Amazon Prime;HBO Max', 'ESPN+, Peacock, Paramount+', 'Netflix', 'Netflix;Hulu;Amazon Prime;Disney+', 'None', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Amazon Prime;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Youtube', 'Netflix;Amazon Prime;HBO Max', 'Amazon Prime', 'Hulu', 'Netflix;Amazon Prime;Disney+', 'Netflix;Amazon Prime', 'Netflix;HBO Max', 'Netflix;HBO Max;Disney+', 'Netflix;HBO Max', 'Disney+', 'Netflix;Hulu;HBO Max', 'Netflix', 'Netflix', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Amazon Prime', 'Netflix;Hulu;Amazon Prime;HBO Max', 'Netflix;Amazon Prime', 'Netflix;Hulu;HBO Max;Disney+', 'none', 'Netflix', 'Netflix;Hulu', 'Netflix;Hulu', 'Netflix;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max', 'Netflix;HBO Max', 'Netflix;Hulu;Amazon Prime;HBO Max', 'Netflix', 'Netflix;HBO Max', 'Netflix', 'Netflix;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Amazon Prime;HBO Max', 'Netflix', 'Netflix;Hulu;HBO Max', 'Netflix;HBO Max', 'Amazon Prime', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime', 'Hulu;Peacock', 'Amazon Prime', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix')
```


```{r}
streaming_platform_section2<-c('Netflix;Hulu;Amazon Prime;Disney+', 'Netflix;Amazon Prime;HBO Max;Disney+', 'Amazon Prime;HBO Max', 'Netflix;Hulu', 'Netflix;Hulu', 'Netflix', 'Free Online Sites', 'Netflix;Hulu', 'Netflix;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Disney+', 'Netflix', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;HBO Max', 'Netflix;Amazon Prime', 'Netflix;Hulu', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;HBO Max', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime;HBO Max', 'Netflix;Amazon Prime;Disney+', 'Netflix;HBO Max', 'Netflix;Hulu;HBO Max', 'Netflix;Disney+', 'Netflix', 'Netflix;Hulu;Amazon Prime;HBO Max', 'Netflix', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Disney+', 'Netflix;Hulu;HBO Max', 'Netflix;Amazon Prime', 'Netflix;Disney+', 'Netflix;Hulu;HBO Max', 'Netflix;Hulu;HBO Max;Disney+', 'Netflix;Hulu;Amazon Prime', 'Netflix;Hulu', 'Netflix;Hulu;Amazon Prime;HBO Max;Disney+', 'Netflix;Hulu', 'Netflix;Hulu;Amazon Prime;HBO Max', 'Netflix', 'Animepahe.com ', 'Netflix;Hulu;HBO Max', 'Netflix;Amazon Prime;HBO Max;hotstar')

```

### Streaming platforms of all students
Concatenate the atomic vectors `streaming_platform_section1`  and `streaming_platform_section2` to obtain a vector named as `streaming_platforms` that consists of the streaming platforms of all students.

*(2 points)*

### Number of unique combinations of streaming platforms
What is the number of unique combinations of streaming platforms used by students?

**Hint:** Use the functions `unique()` and `length()`

*(3 points)*

### Most popular combination
What is the most popular combination of streaming platforms used by students? How many students use it?

**Hint:** Use the function `table()`

*(4 points)*

### Frequency of a streaming platform
Write a function that takes a streaming platform *(for example, 'Netflix')* as an argument, and returns the number of students using that platform.

Call the function to find the number of students using:

#### Netflix
#### Hulu

Note that a student using a particular platform may be using other platforms as well. 

*(8 points)*

### At least 3 platforms
How many students use at least 3 streaming platforms?

*(6 points)*

## Prime Factors with and without `for` loops

### Prime number
Define a function named as `prime` that checks if an integer is prime. The function must accept the integer as an argument, and return `TRUE` if the integer is prime, otherwise it must return `FALSE`.

Call your function with the argument as 197.

*(2 points)*

### Factor

Define a function named as `factor` that checks if an integer is a factor of another integer. The function must accept both the integers as arguments, and return `TRUE` if the integer is a factor, otherwise it must return `FALSE`.

Call your function with the arguments as (19,85)

*(2 points)*

### Prime factors

Define a function named as `prime_factors` that returns the prime factors of an integer. The function must use the functions `prime` and `factor`.

Call the function with the argument as `1234567`, and print the returned object, which should be the prime factors of the number.

*(4 points)*

### Prime factors code: Time

Record and print the time taken to execute the function `prime_factors` for the integer `1234567`.

*(3 points)*

### Prime number: Without `for` loop
Update the function `prime` so that it works without a `for` loop. Name the updated function as `prime_nofor`. Call the function `prime_nofor` with the argument as 197.

*(6 points)*

### Prime factors: Without `for` loop
Update the function `prime_factors` so that it works without a `for` loop. Name the updated function as `prime_factors_nofor`. The function must use the functions `prime_nofor` and `factor`. Call the function `prime_factors_nofor` with the argument as `1234567`, and print the returned object, which should be the prime factors of the number.

*(12 points)*

### Prime factors: Time taken without `for` loop

Record and print the time taken to execute the function `prime_factors_nofor` for the integer `1234567`. Is it less than the time taken by `prime_factors` for the integer `1234567`? 

*(3 points)*

## Air quality sensors

Air quality sensors are used to measure the amount of contaminants in air. This question will guide you in finding the location of installing 50 air quality sensors in the State of Colorado, such that they are as far away from each other as possible. The approach below is a greedy algorithm to find an approximate [Maximin design](https://www.sciencedirect.com/science/article/abs/pii/037837589090122B).

Use the following code to generate the coordinate-pairs (latitude and longitude) of potential locations for installing  an air quality sensor in Colorado:

```{r}
x1<-seq(37,41,length=100)
x2<-seq(102,109,length=100)
candidate_set<-as.matrix(expand.grid(x1,x2))
```

### Number of coordinate-pairs
How many coordinate-pairs are there in the object `candidate_set`?

Note that:

1. A coordinate-pair means a latitude-longitude pair.

2. 'Air quality sensor' will be referred as 'sensor' in the questions below for brevity.

*(2 points)*

### First sensor
The first sensor is to be installed closest to Denver *(closest in terms of Euclidean distance)*. Find the coordinate-pair of the location where the first sensor will be installed. The coordinate-pair of Denver is: [39.7392$^{\circ}$ N, 104.9903$^{\circ}$ W] 

Note that the suffixes $^{\circ}$ N and $^{\circ}$ W are omitted in the object `candidate_set`.

*(6 points)*

### Second sensor
Find the coordinate-pair of the installation of the next sensor, such that it is as far as possible from the first sensor installed near Denver.

*(6 points)*

### Air sensor coordinates
Use the `rbind()` function to stack the coordinate-pairs of the first and second sensors vertically to obtain a 2 x 2 matrix. Name the matrix as `air_sensor_coordinates`. 

Run the code below to check if your results seem correct. The coordinate-pairs of the two air quality sensors will be marked as black dots over the map of Colorado. Make sure the file *colorado.png* is in the same directory as the *qmd file.

```{r}
#| eval: false

library(png)
if(!require(png)){
  install.packages("png")
  library(png)
}
library(png)
my_image <- readPNG("colorado.png")
plot(-x2,x1, type='n', main="", xlab="longitude", ylab="latitude")
rasterImage(my_image, xleft=-109, xright=-102, ybottom=37, ytop=41)
points(-air_sensor_coordinates[,2],air_sensor_coordinates[,1],pch=16)

```

*(4 points)*

### Third sensor
Now you need to find the coordinate-pair for installing the third sensor such that it is far away from the two already-installed sensors. Proceed as follows:

1. Find the minimum distance of each coordinate-pair in `candidate_set` from the two already installed sensors. For example, if a coordinate-pair is at a distance of 5 units from the first sensor, and 10 units from the second sensor, then its minimum distance from the sensors will be $\min(5,10) = 5$ units. 

2. Select the coordinate-pair (from `candidate_set`) whose minimum distance from the two already installed sensors is the maximum. 

3. Stack the coordinate-pair of the third air quality sensor vertically on the array `air_sensor_coordinates`. 

Execute the code provided in the previous question to check if your results seem correct. The coordinate-pairs of the three air quality sensors will be marked as black dots over the map of Colorado.

**Hint:**

For step (1) above:

1. Define a function which computes the distances of a coordinate-pair from all the coordinates of `air_sensor_coordinates`, and returns the minimum distance.

2. Apply the function on all the coordinate-pairs in `candidate_set` using the `apply()` function.

*(16 points)*

### 47 more sensors

You need to find 47 more coordinate-pairs to install air quality sensors well-spread across Colorado. We will generalize the steps in the previous question as follows:

1. Suppose you have already found the coordinate-pairs for the installation of *i* sensors.


2. Find the minimum distance of each coordinate in `candidate_set` from the *i* already installed sensors. For example, if a coordinate-pair is at a distance of $d_1$ from the first sensor, $d_2$ from the second sensor,..., and $d_i$ from the $i^{th}$ sensor, then its minimum distance from the sensors will be $min(d_1, d_2, ..., d_i$). 


3. Select the $i+1^{th}$ coordinate-pair (from `candidate_set`) as the one whose minimum distance from the $i$ already installed sensors is the maximum. 
Execute the code provided in the previous question to check if your results seem correct. You should see 50 black dots well spread across Colorado.

*(6 points)*