---
title: "R: Recursion"
format:
    html:
      toc: true
      self-contained: true
editor_options: 
  chunk_output_type: console
---

## Recursion

Recursion is a method of solving a problem by dividing it into smaller instances of the same problem. Recursion solves such problems by using functions that call themselves from within their own code.

Below is an example, where we defined a function that computes the factorial of a number by recursion.

```{r}
factorial<-function(n)
{
  if(n==1){return(1)}else(return(n*factorial(n-1)))
}
factorial(5)
```

### Practice exercise 1

Use recursion to write a function that accepts a word as an argument, and returns `TRUE` if the word is a palindrome, otherwise returns `FALSE`.


```{r}
#| echo: false
#| eval: false

word<-'reduyder'
palindrome<-function(word)
{
  if(nchar(word)<=1)
  {
    return(TRUE)
  }else if(substr(word,1,1)==substr(word,nchar(word),nchar(word)))
  {
    palindrome(substr(word,2,nchar(word)-1))
  }else{
    return(FALSE)
  }
}
palindrome(word)
```