---
title: "R: Objects"
format:
    html:
      toc: true
      self-contained: true
editor_options: 
  chunk_output_type: console
---
  
## Atomic vectors
  
An atomic vector in R is a vector containing objects of the same datatype. If the objects are not of the same datatype, then they are coerced to be of the same datatype. It is defined using the keyword `c()`.

```{r}
numbers = c(1,2,67)
```

The in-built R function `length()` is used to find the length of an atomic vector.

```{r}
length(numbers)
```

### Slicing the atomic vector

An atomic vector can be sliced using the indices of the elements within  `[]` brackets.

For example, consider the vector:

```{r}
vec<-1:40
```

Suppose, we wish to get the $3^{rd}$ element of the vector. We can get it using the index `3`:

```{r}
vec[3]
```

A sequence of consecutive elements can be sliced using the indices of the first element and the last element around the `:` operator. For example, let us slice elements from the $3^{rd}$ index to the $10^{th}$ element of the vector `vec`:

```{r}
vec[3:10]
```

We can slice elements at different indices by putting the indices in an atomic vector within the `[]` brackets. Let us slice the $4^{th}$, $7^{th}$, and $18^{th}$ elements of the vector `vec`:

```{r}
vec[c(4,7,18)]
```

We can slice consecutive elements, and non-consecutive elements simultaneously. Let us slice the elements from the $4^{th}$ index to the $9^{th}$ index and the $30^{th}$ and $36^{th}$ element.

```{r}
vec[c(4:9,30,36)]
```

### Removing elements from atomic vector

Elements can be removed from the vector using the negative sign within `[]` brackets.

Remove the 2nd element from the vector:

```{r}
vec<-1:5
vec[-2]
```

If multiple elements need to be removed, the indices of the elements to be removed can be given as an atomic vector.

Remove elements `2` to `6` and element 10 from the vector:
```{r}
vec<-1:20
vec[-c(2:6,10)]
```


**Example:** USA’s GDP per capita from 1960 to 2021 is given by the vector `G` in the code chunk below. The values are arranged in ascending order of the year, i.e., the first value is for 1960, the second value is for 1961, and so on. Store the years in which the GDP per capita of the US increased by more than 10%, in a vector.

```{r}
G = c(3007, 3067, 3244, 3375,3574, 3828, 4146, 4336, 4696, 5032,5234,5609,6094,6726,7226,7801,8592,9453,10565,11674,12575,13976,14434,15544,17121,18237,19071,20039,21417,22857,23889,24342,25419,26387,27695,28691,29968,31459,32854,34515,36330,37134,37998,39490,41725,44123,46302,48050,48570,47195,48651,50066,51784,53291,55124,56763,57867,59915,62805,65095,63028,69288)
```

```{r}
years<-c()
for(i in 1:(length(G)-1))
{
  diff = (G[i+1]-G[i])/G[i]
  if(diff>0.1){years<-c(years,1960+i)}
}
print(years)
```

## Matrix

Matrices are two-dimensional arrays. The in-built function `matrix()` is used to define a matrix. An atomic vector can be organized as a matrix by specifying the number of rows and columns.

For example, let us define a 2x3 matrix (2 rows and 3 columns) consisting of consecutive integers fro1 1 to 6.

```{r}
mat<-matrix(1:6,2,3)
mat
```

Note that the integers fill up column-wise in the matrix. If we wish to fill-up the matrix by row, we can use the `byrow` argument.

```{r}
mat<-matrix(1:6,2,3, byrow = TRUE)
mat
```

The functions `nrow()` and `ncol()` can be used to get the number of rows and columns of the matrix respectively.

```{r}
nrow(mat)
```

```{r}
ncol(mat)
```

Matrices can be sliced using the indices of row and column separated by a `,` in box brackets. Suppose we wish to get the element in the $2^{nd}$ row and $3^{rd}$ column of the matrix:

```{r}
mat[2,3]
```

For selecting all rows or columns of a matrix, the index for the row/column can be left blank. Suppose we wish to get all the elements of the $1^{st}$ of the matrix:

```{r}
mat[1,]
```

Row and columns of the matrix can be sliced using the `:` operator. Suppose we want to select a sub-matrix that has elements in the first two rows and columns 2 and 3 of the matrix `mat`:

```{r}
mat[1:2,2:3]
```

Suppose we need to sum up all the rows of the matrix. We can do it using a `for` loop as follows:

```{r}
row_sum<-c(0,0)
for(i in 1:nrow(mat))
{
  for(j in 1:ncol(mat))
  {
    row_sum[i]<-row_sum[i]+mat[i,j]
  }
}
row_sum
```

Observe that in the above `for` loop, elements of each row are added one at a time. We can add all the elements of a row simultaneously using the `sum()` function. This will reduce a `for` loop from the above code:

```{r}
row_sum<-c(0,0)
for(i in 1:nrow(mat))
{
    row_sum[i]<-sum(mat[i,])
}
row_sum
```


In the above code, we sum up all the elements of the row simultaneously. However, we still need to sum up the elements of each row one at a time.

### The `apply()` function

The `apply()` function can used to apply a function simultaneously on all rows or columns of a matrix. Thus, this function helps avoid a `for` loop to iterate over all the rows and columns of the matrix. This reduces the execution time of the code since operations are performed in-parallel, instead of one-at-a-time in a `for` loop.

Let us use the `apply()` function to sum up all the rows of the matrix `mat`.

```{r}
apply(mat,1,sum)
```

Let us compare the time taken to sum up rows of a matrix using a `for` loop with the time taken using the `apply()` function. 

```{r}
options(digits.secs = 6)
start.time <- Sys.time()
row_sum<-c(0,0)
for(i in 1:nrow(mat))
{
    row_sum[i]<-sum(mat[i,])
}
row_sum
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```


```{r}
start.time <- Sys.time()
apply(mat,1,sum)
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```

Observe that the `apply()` function takes much lesser time to sum up all the rows of the matrix as compared to the `for` loop.

Recall the earlier example where we computed year's in which the increase in GDP per capita was more than 10%. Let us use matrices to solve the problem. We'll also compare the time it takes using a matrix with the time it takes using `for` loops.

```{r}
start.time <- Sys.time()

#Let the first column of the matrix be the GDP of all the years except 1960, and the second column be the GDP of all the years except 2021.
GDP_mat<-matrix(c(G[-1],G[-length(G)]),length(G)-1,2)

#The percent increase in GDP can be computed by performing computations using the 2 columns of the matrix
inc<-(GDP_mat[,1]-GDP_mat[,2])/GDP_mat[,2]
years<-1961:2021
years<-years[inc>0.1]
years
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```


Without matrices, the time taken to perform the same computation is  measured with the code below.

```{r}
start.time <- Sys.time()
years<-c()
for(i in 1:(length(G)-1))
{
  diff = (G[i+1]-G[i])/G[i]
  if(diff>0.1){years<-c(years,1960+i)}
}
print(years)
#print(proc.time()[3]-start_time)
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```

Observe that matrices reduce the execution time of the code as computations are performed simultaneously, in contrast to a `for` loop where computations are performed one at a time.

