---
title: "R: Objects"
format:
    html:
      toc: true
      self-contained: true
editor_options: 
  chunk_output_type: console
---
  
## Atomic vectors
  
An atomic vector in R is a vector containing objects of the same datatype. If the objects are not of the same datatype, then they are coerced to be of the same datatype. It is defined using the keyword `c()`.

```{r}
numbers = c(1,2,67)
```

The in-built R function `length()` is used to find the length of an atomic vector.

```{r}
length(numbers)
```

### Removing elements from atomic vector

Elements can be removed from the vector using the negative sign within `[]` brackets.

Remove the 2nd element from the vector:

```{r}
vec<-1:5
vec[-2]
```

If multiple elements need to be removed, the indices of the elements to be removed can be given as an atomic vector.

Remove elements `2` to `6` and element 10 from the vector:
```{r}
vec<-1:20
vec[-c(2:6,10)]
```

**Example:** USA’s GDP per capita from 1960 to 2021 is given by the vector `G` in the code chunk below. The values are arranged in ascending order of the year, i.e., the first value is for 1960, the second value is for 1961, and so on. Store the years in which the GDP per capita of the US increased by more than 10%, in a vector.

```{r}
G = c(3007, 3067, 3244, 3375,3574, 3828, 4146, 4336, 4696, 5032,5234,5609,6094,6726,7226,7801,8592,9453,10565,11674,12575,13976,14434,15544,17121,18237,19071,20039,21417,22857,23889,24342,25419,26387,27695,28691,29968,31459,32854,34515,36330,37134,37998,39490,41725,44123,46302,48050,48570,47195,48651,50066,51784,53291,55124,56763,57867,59915,62805,65095,63028,69288)
```

```{r}
options(digits.secs = 6)
start.time <- Sys.time()

years<-c()
for(i in 1:(length(G)-1))
{
  diff = (G[i+1]-G[i])/G[i]
  if(diff>0.1){years<-c(years,1960+i)}
}
print(years)
#print(proc.time()[3]-start_time)
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```

## Matrix

Matrices are two-dimensional arrays. The in-built function `matrix()` is used to define a matrix. An atomic vector can be organized as a matrix by specifying the number of rows and columns.

For example, let us define a 2x3 matrix (2 rows and 3 columns) consisting of consecutive integers fro1 1 to 6.

```{r}
matrix(1:6,2,3)
```

Note that the integers fill up column-wise in the matrix. If we wish to fill-up the matrix by row, we can use the `byrow` argument.

```{r}
matrix(1:6,2,3, byrow = TRUE)
```

The functions `nrow()` and `ncol()` can be used to get the number of rows and columns of the matrix respectively.

```{r}
nrow(mat)
```

```{r}
ncol(mat)
```

Matrices can be sliced using the indices of row and column separated by a `,` in box brackets. Suppose we wish to get the element in the $2^{nd}$ row and $3^{rd}$ column of the matrix:

```{r}
mat[2,3]
```

For selecting all rows or columns of a matrix, the index for the row/column can be left blank. Suppose we wish to get all the elements of the $1^{st}$ of the matrix:

```{r}
mat[1,]
```

Row and columns of the matrix can be sliced using the `:` operator. Suppose we want to select a sub-matrix that has elements in the first two rows and columns 2 and 3 of the matrix `mat`:

```{r}
mat[1:2,2:3]
```

Suppose we need to sum up all the rows of the matrix. We can do it using a `for` loop as follows:

```{r}
mat<-matrix(1:6,2,3)
row_sum<-c(0,0)
for(i in 1:nrow(mat))
{
  for(j in 1:ncol(mat))
  {
    row_sum[i]<-row_sum[i]+mat[i,j]
  }
}
row_sum
```

Observe that in the above `for` loop, elements of each row are added one at a time. We can add all the elements of a row simultaneously using the `sum()` function. This will reduce a `for` loop from the above code:

```{r}
mat<-matrix(1:6,2,3)
row_sum<-c(0,0)
for(i in 1:nrow(mat))
{
    row_sum[i]<-sum(mat[i,])
}
row_sum
```


In the above code, we sum up all the elements of the row simultaneously.



A very useful application of matrices is that parallel computations can be performed on rows or columns of a matrix using the `apply()` function.

